Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for

### 2020-09-02 Scripts/Manuscript.Rmd
Drafted intrto, methods and results

---
title: Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for
  Policy Response Under Climate Change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, U. Rashid Sumaila^1^,  Vicky
  Lam^1^, David VanderZwaag^3^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British
  Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute,
  University of Bern. ^3^Schulich School of Law, Dalhousie University'
output:
  word_document: default
  pdf_document: default
subtitle: Climate-ready transboundary fisheries management under Climate Change
---

@@

The delineation of Economic Exclusive Zones (EEZs) by the United Nations Convention on the Law of the Seas (UNCLOS) in the early 80s, established political boundaries across the distribution of many fish species, creating shared species between nations. Despite not implicit, UNCLOS incentivizes nations to cooperate in the management of transboundary fisheries and most management plans emplaced are based on historical distribution of catches. Climate change represents a challenge for the joint management of fish stocks. For instance, climate-driven changes in species distributions have led to conflict between nations due to changes in the proportion of captures, quota allocation, and fishery newcomers. Recent events suggests that even changes in species that contribute a relatively small proportion of the overall catch can lead to important international conflict when no legal framework is available. Thus, it is fundamental to prepare for climate change-driven impacts such as identifying the stocks that are going to change historical proportions within neighboring EEZs, the degree by which they will change, and the time when such change is going to happen. Only by adapting management plans and policy that anticipates to these effects will international fisheries treaties be climate-ready. Here, we build on previous work to identify the transboundary stocks that will see changes in their stock-share-proportion (e.g. the proportion of the stock that each EEZ has) around the world. Moreover, we determine the time of emergence of each fish stock, defined as the year in which the stock-share-proportion will change from its historical range. Our results suggest that by the mid of the 21st century EEZs will see changes in stock-share ratio between 1 and 10% with some exceeding 25%. Most of these changes will happen before 2050 urging policy to be re-formulated to account for the possible implications of these changes. By the end of the 21st century virtually all EEZs and transboundary species will have experienced changes in stock-share-proportion relative to historical distributions. The results of this research has the potential to improve climate-ready fisheries if policy is modify to respond to the consequences presented here.

<!-- # Introduction -->

The world is moving towards sustainable development and ocean-based activities plays a mayor role in achieving international policies.

One of the main goals of the United Nations Sustainable Development Goal No.14 (SDG 14-Life below water) is to achieve sustainable development of fisheries [@SDG14].
<!-- (https://sustainabledevelopment.un.org/sdg14). -->

Climate change is expected to reduce the production from global fisheries in about XX% to xx% [@Cheung; @Barange] compromising our capacity to reach SDG 14.

Recent research show that the impacts of climate change on the productivity of marine wild fisheries can be offset with the proper reform of current fisheries management [@Gaines; @CheungClimateReadyFish]. While bringing fisheries that are poorly managed to a sustainable level will likely offset the negative impacts of climate change to fisheries production [@Gaines], it will not necessarily solve the spatial consequences of species moving from one management area to another.

Shifts in species distribution does not necessarily mean a reduction in production (e.g. reduction in stock biomass) but rather a spatial re-distribution of the stock's biomass [@RefLiam?].

Documented cases of stock shifts have resulted in international conflict and sustainable fishing rates by parties involved in the conflict.

This is of specific concern to a set of 633 fished species that cross international borders, also known as transponder species [@FishForVisa].

With the implementation of arbitrary spatial management boundaries, at least 633 species became transboundary representing an average catch (2005 - 2010) of 47.6 million tons and generating USD 65,483 million in fishing revenue [@FishorVisa].

Transboundary fish stocks refer to those species that cross international jurisdictions known as Economic Exclusive Zones (EEZs) of neighboring nations [@FAO]. These stocks are responsible for an average catch (2005 - 2010) of 47.6 million tons generating USD 65,483 million in fishing revenue [@FishorVisa].

Historically, environmentally-driven shits in shared stocks have cause problems and threaten stock's sustainability. **In early 1980's Canada and the US blablabla.** Many fish stocks managed by the European Union ICES regultaroy areas have been shifting since the 1970's [@EuropeShift]. Such shifts have already result in confrontations between nations as evidenced by the expansion of the Atlantic mackerel (*Scomber scombrus*) distribution reaching the EEZ of Iceland [@Spijkers:2017ij].  More recently, Chile and Peru have declare a cooperation strategy to mange the transboundary stock of Peruvean Anchovera [@Peru]. Furthermore, as we continue to enter the $21^{st}$ century, commercial marine species are expected to continue shifting (mostly poleward) potentilly compromising shared stocks management. By the end of the century, all EEZs of the world are proyected to have at least one new transboundary species with some of them having over 10 new stocks [@Pinsky]. Consequently, as species occupy new waters they leave historic areas and so, xx% of species are expected to exit EEZs by the end of che century [@Oremus]. This entering and exicting national jurisdicions will complicate fisheries management potentially creating confllict between neighbouring nations [@Pinksy @Oremus:2020id].

 So, preparing policy to species on the move is key to achieve sustainable development goals and proper governance of the world fisheries [@Pinksy].

Undrstanding the degree by which shared stocks are going to change and the time of emergence is key for creating climate-ready fisheries policy.

The concepts of transboundary stocks and EEZs were introduced in the early 1980's by the United Nations Law of the Sea (UCLOS) [@UnitedNations:1986tl]. Under UNCLOS each costal nation has the right (and responsabillity) to manage wild fish stocks up to 200 nautical mailes of the coast. Exceptions exist when territorial waters are in dispute (e.g. South China sea[@Ref]) or when menagement agreements exist between nations [@Ref]. In fact, UNCLOS incentives actions to cooperate on the management of shared stocks [Article 63, @UnitedNations:1986tl]. Climate change is shifting the distribution of marine species to the extreem that some spatial policies implemented to manage fisheries are becoming obsolete [@Pololanza; @Jessica].

<!-- ## Link paper un terms of SDGs -->
<!-- - UnitedNations2018 Blanchard:2017hl Singh:2017ds -->

<!-- ## Collaborative management of share stocks -->
<!-- - Theat points -->
<!-- - Concept of treaty estabillity -->

@@

The Food and Agricultural Organization (FAO) of the UN define transboundary stocks as those that freely swim across the EEZ of neighboring nations [@REFFAO]. Accordingly, we included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs [@Palacios-Abrantes:xxxx]. We consider each species as a single stock unit within an EEZ due to a lack of more detailed spatial information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the *Sea Around Us* catch data to estimated the average catch and fishing revenue (USD 2019). We only used data from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we catalogue catch trends as increasing, constant and decreasing, following Palacios-Abrates *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region [@UN] and biome [@Gabriel], and species were attributed a habitat preference [@FishBase]. For the present study we first projected the 633 species distributions under climate change and then estimated two main metrics; a stock's share ratio (SSR) and time of emergence (ToE) as described in the following sections.

@@

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in an unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members [@Rodggers:2015jq].

@@

In game theory, a threat point is known as the minimum payoff that a player is willing to receive in order to cooperate with other players. Thus, if a player's payoff is lower than the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@Sumaila:E&S]. Here, we adopted the concept of threat point as the minimum proportion of the catch that each country will accept if they were to cooperate in the management of a transboundary stock [@Palacios-Abrantes:E&S] hereafter referred as stock share ratio (SSR). For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@Palacios-Abrantes:xxxx]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^st$ century as the average of 2021 to 2040 ($t_e$), and mid $21^st$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. We then estimate the percentage change in the SSR of each future time period relative to the historic time period. To address the natural spatial variability of the stock we only considered a SSR change when the future SSR was higher or lower than the average historic SSR $\pm$ 1 s.d. The analysis was replicated for each of the ten ensemble members capturing variation by averaging results among ensemble members (mean $\pm$ standard deviation).

## Time of Emergence Index

Knowing the time (e.g. year) in which SSR is going to diverge from a historical variability can be key for policy to prepare to such change. In here we call this event time of emergence and defined it as the moment in time when the future variation in SSR becomes larger than the historical trend variability [@Roggers:2015jq]. We employed the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupy. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated following the previous equation.

@@

In other words, for two neighboring EEZs sharing a stock, if the centroid shifts towards EEZ A ($D_A$ decreases) but away from EEZ A B ($D_B$ increases), EEZ A may catch a bigger share over B and the risk of dispute increases (see *TI* equation). Consequently, if the centroid shifts are within historical variability (*SNR* < 1), risk of dispute between EEZ A and B is lower, finally, if the centroid shifts are beyond historical variability (*SNR* > 1), the risk of dispute is higher. Thus, risk is positively related to the frequency of centroid shifts beyond historical variability.

@@

We tested data for normallity (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the time of emergence of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we lookd at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the time of emergence of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@], *cowplot* [@], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated changes in the Stock Share Ratio of transboundary species shared by neighboring EEZs by early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). While most EEZs will experience an average change lower than 25%, a few EEZs are expected to have larger changes in SSR relative to early century. Such change is expected to slightly increase in intensity towards the  mid $21^{st}$ century (**Supplements Fig 1 (histogram)**). For example, Guatemala's SSR of *Panulirus gracilis* shared with Mexico is projected to go from 12% to 26%. The number of stocks per EEZ that will see a change in SSR by early $21^{st}$ century is relatively low (median of 13 spp per EEZ), however, some regions like West Africa and the Mediterranean will see changes in over 50 species (e.g. Senegal, 94). These regions, where EEZs are relatively small and have multiple borders will see a complex dynamic of changes in SSR with no clear patterns (Fig. 1B). Towards the mid $21^st$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to early century, but are expected to increase in intensity (**Supplements Fig 1 (histogram)**).

@@

![Average gain of transboundary stock share ratio (lines) and number of shifting stocks (polygon) by 2030 (2021-2040) relative to 1951-2005. Arrows go from the losing EEZ (point) to the winner (arrowhead). Pannel B zooms in to specific areas ](../Figures/Fig1.png)

@@

We estimated the ToE between neighboring EEZs for 633 transboundary stocks, that is, the first year where the future stock's distrtibution overpasses the historical variation. Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). The cumulative sum of emerging stocks will have a steeper increase until 2040 where the increment will be slightly less intense (Fig. 2). Similarly, for the number of EEZs, the increment curve seem to flatten by 2020. By 2081, all of the EEZs analyzed here will have at least one emerging stock while 2097 will be the year where all 633 stocks have emerged in at least one neighboring EEZ (Fig. 2).

@@

![Cumulative increase in time of emergence by EEZ and stock](../Figures/Fig2.png)

On average, most of the world EEZ's will see a stock emerge by 2030 with a few EEZs having a relative late average ToE (e.g. Honduras average ToE is 2071) and 54 EEZs having already emergered (e.g. average ToE >= 2020). A higher number of emerging stocks will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Carribeean and the Medditerranean) (Fig 2). In most cases, the current participation of the emerging stocks in a country's total fishing revenue is less than 50%. However, in some cases emerging stocks represent 85% of total fishing revenue (e.g. Peru) or more (e.g. Marshall Islands, 92%). Only in very few cases like Brazil and Alaska's polar EEZ, presented no emerging stock between 2006 and 2100.

@@

![(Average time of emergence (ToE) per fishing entity. Fishing revenue in logarithmic sacle. Showing names for the top 75th percentile). Countries marked with a asterisk have at least one border](../Figures/Fig4.png)

>>>>

---
title: Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for
  Policy Response Under Climate Change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2^, U. Rashid Sumaila^1^,  Vicky Lam^1^, David VanderZwaag^3^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British
  Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute,
  University of Bern. ^3^Schulich School of Law, Dalhousie University'
output:
  word_document: default
  pdf_document: default
---

@@

<!-- The delineation of Economic Exclusive Zones (EEZs) by the United Nations Convention on the Law of the Seas (UNCLOS) in the early 80s, established political boundaries across the distribution of many fish species, creating shared species between nations. Despite not implicit, UNCLOS incentivizes nations to cooperate in the management of transboundary fisheries and most management plans emplaced are based on historical distribution of catches. Climate change represents a challenge for the joint management of fish stocks. For instance, climate-driven changes in species distributions have led to conflict between nations due to changes in the proportion of captures, quota allocation, and fishery newcomers. Recent events suggests that even changes in species that contribute a relatively small proportion of the overall catch can lead to important international conflict when no legal framework is available. Thus, it is fundamental to prepare for climate change-driven impacts such as identifying the stocks that are going to change historical proportions within neighboring EEZs, the degree by which they will change, and the time when such change is going to happen. Only by adapting management plans and policy that anticipates to these effects will international fisheries treaties be climate-ready. Here, we build on previous work to identify the transboundary stocks that will see changes in their stock-share-proportion (e.g. the proportion of the stock that each EEZ has) around the world. Moreover, we determine the time of emergence of each fish stock, defined as the year in which the stock-share-proportion will change from its historical range. Our results suggest that by the mid of the 21st century EEZs will see changes in stock-share ratio between 1 and 10% with some exceeding 25%. Most of these changes will happen before 2050 urging policy to be re-formulated to account for the possible implications of these changes. By the end of the 21st century virtually all EEZs and transboundary species will have experienced changes in stock-share-proportion relative to historical distributions. The results of this research has the potential to improve climate-ready fisheries if policy is modify to respond to the consequences presented here. -->

# Introduction

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, increasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species are changing their distribution [@ Poloczanska:2016kk] and will continue to do so in the search of the apropiate environmental conditions needed for their subsistence [@Cheung:2010dt. Climate change is expected to impact fisheries production at different latitudes compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@SDG14; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be offset with the proper reform of current fisheries management [@Gaines:2018sg; @CheungClimateReadyFish]. However, such reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg;@Pinsky:2018cb;].

The concept of shared stocks gained attention in the early 1980's when the United Nations officialized the concept of Exclusive Economic Zone (EEZs) though the Law of the Sea (UCLOS) [@UnitedNations:1986tl]. The implementation of arbitrary jurisdictions across the global ocean created a partition in the distribution of several stocks, thus, creating what we known today as shared stocks. As defined by the FAO, shared stocks can be classified as transboundary, those that cross neighboring EEZs; straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and highly migratory, mainly tunas and bill-fishes that cover broad oceanic regions [@FAO; @Munro]. While counties are responsible for the management of stocks within their EEZs, UNCLOS incentives actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fished stocks are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 [@FishorVisa].

The management, and thus sustainability, of fisheries that target shared stocks is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. There is evidence that such shifts have already complicated shared marine resources around the globe. In the North-eastern Atlantic, evidence suggest that shared stocks managed within the European Union divisions have been shifting since the 1970's compromising the region's sustainability [@EuropeShift]. In the North-eastern Pacific, sockeye salmon (*Oncorhynchus nerka*) shifted in the 1980's towards Alaska, creating a ten years conflict between Canada and the United States [@Song]. In the South-eastern Pacific, the southern stock of Peruvean anchoveta (*Engraulis ringens*) shared by Chile and Peru has gradually shifted south (towards Chilean waters) adding to the over-exploitation of the resource [@Cashion:2018cg]. Global shifts in the distribution of tunas [@ErauskinExtramiana:2019fw] have brought additional management challenges in both the Western Indian ocean and South-western Pacific [@Miller:2007kx]. Moreover, as species continue shifting with climate change, current policies might become obsolete, and thus, require new management strategies that create equitable systems across all dependent fishing nations [@Bell:2013ga; @Pinsky:2018cb]. We argue that two concepts will be essential in addressing the challenges that climate change will bring to the management of transboundary fisheries.

The first concept alludes to the game theory concept of thread point, known as the minimum payoff a player is willing to receive in order to cooperate with others. Thus, if a player's future payoff drops below the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@Sumaila:E&S].This can be seen as the proportion of a species distribution that each neighboring nation has (or the quota allocation that each one gets) at the time of signing off an agreement [@PalaciosAbrantes:E&S]. Many shared fisheries operate this way [@EuropeShift; @GoMA]. If not addressed, future changes in the thread point (hereafter called Stock Share Ratio -SSR) of jointly managed fisheries can have adverse consequences for shared fisheries [@PalaciosAbrantes:E&S; @Sumaila:E&S]. The second concept refers to the Time of Emergence (ToE), defined as the moment when a future trend of a particular variable exceeds the historical variability [@IPCC:2019to]. Such concept has been used to look at the emergence of diverse oceanic variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg] and ecological consequences of marine heatwaves [@Cheung:2020ja]. Similarly, climate change is expected to drive biomass further away from the historic range limits of a species [@Alexa?]. Assessing the ToE of shared stocks is as important as the change itself as it could reduce potential international conflicts [@Spijkers:2017ij] and prepare for emerging fisheries [@Pinsky:2018cb;  @Pinsky:2014by].

Here we estimate the changes in Stock Share Ratio and the Time of Emergence of 633 transboundary stocks shared by 280 EEZs of 198 coastal countries under climate change (See methods). We rely on a species distribution and Earth system model with ten ensemble members to project the distribution of the 663 selected fish stocks under a high greenhouse gas emission scenario. We treated each species in an EEZ as a single stock due to a lack of more detailed spatial data [@FishForVisa] and only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g. we did not include the high seas). Our results suggest that virtually all EEZs will have a change in SSR by 2030 and that 87% of them already had at least one emerging species since 2006, with significant regional and environmental differences. Towards the end of the $21^{st}$ century, all of the 280 EEZs will have at least one emerging stock and all 633 stocks would have emerged at least once. Our results have direct implications for marine governance of shared marine resources under a changing world.

@@

The Food and Agricultural Organization (FAO) of the UN define transboundary stocks as those that freely swim across the EEZ of neighboring nations [@REFFAO]. Accordingly, we included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs [@Palacios-Abrantes:xxxx]. We consider each species as a single stock unit within an EEZ due to a lack of more detailed spatial information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the Sea Around Us data to estimate the catch and fishing revenue (USD 2019 from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we classified catch trends as increasing, constant and decreasing, following Palacios-Abrantes *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region [@UN] and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase].

@@

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members [@Rodggers:2015jq].

@@

We adopted the concept of threat point as the minimum proportion of the stocks’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@Palacios-Abrantes:E&S]. For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@Palacios-Abrantes:xxxx]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. We then estimate the percentage change in the SSR of each future time period relative to the historic time period. To address the natural spatial variability of the stock we only considered an  SR change when the future SSR was higher or lower than the average historic SSR $\pm$ 1 s.d. The analysis was replicated for each of the ten ensemble members capturing variation by averaging results among ensemble members (mean $\pm$ standard deviation).

Knowing the time (e.g. year) in which a shared stock is going to diverge from a historical variability can be key for preparing to shifting stocks. We employed the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupy. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated following the SNR method.

@@

<!-- In other words, for two neighboring EEZs sharing a stock, if the centroid shifts towards EEZ A ($D_A$ decreases) but away from EEZ A B ($D_B$ increases), EEZ A may catch a bigger share over B and the risk of dispute increases (see *TI* equation). Consequently, if the centroid shifts are within historical variability (*SNR* < 1), risk of dispute between EEZ A and B is lower, finally, if the centroid shifts are beyond historical variability (*SNR* > 1), the risk of dispute is higher. Thus, risk is positively related to the frequency of centroid shifts beyond historical variability. -->

@@

We tested data for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@], *cowplot* [@], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated changes in the SSR of transboundary species shared by neighboring EEZs by early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). While most EEZs will experience an average change lower than 25%, a few EEZs are expected to have larger changes in SSR. Such change is expected to slightly increase in intensity towards the  mid $21^{st}$ century (**Supplements Fig 1 (histogram)**). For example, Guatemala's SSR of *Panulirus gracilis* shared with Mexico is projected to go from 12% to 26%. The number of stocks per EEZ that will see a change in SSR by early $21^{st}$ century is relatively low (median of 13 spp. per EEZ), however, some regions like West Africa and the Mediterranean will see changes in over 50 species (e.g. Senegal, 94). These regions, where EEZs are relatively small and have multiple borders will see a complex dynamic of changes in SSR with no clear patterns (Fig. 1B). Towards the mid $21^st$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to early century, but are expected to increase in intensity (**Supplements Fig 1 (histogram)**).

@@

![Figure 1. Average gain of transboundary stock share ratio (lines) and number of shifting stocks (polygon) by 2030 (2021-2040) relative to 1951-2005. Arrows go from the losing EEZ (point) to the winner (arrowhead). Panel B zooms in to specific areas.](../Figures/Fig1.png)

@@

We estimated the ToE between neighboring EEZs for 633 transboundary stocks, that is, the first year where the future stock's distribution overpasses the historical variation. Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). The cumulative sum of emerging stocks will have a steeper increase until 2025 where the increment will be slightly less intense (Fig. 2). Similarly, for the number of EEZs, the increment curve seem to flatten by 2020. By 2081, all of the EEZs analyzed here will have at least one emerging stock while 2097 will be the year where all 633 stocks have emerged in at least one neighboring EEZ (Fig. 2).

@@

![Figure 2 Cumulative increase in time of emergence by EEZ and stock](../Figures/Fig2.png)

Most of the world EEZ’s will see an average ToE across transboundary stocks by 2030 with a few EEZs having a relative late average ToE (e.g. Honduras average ToE is 2071) and 54 EEZs having already emerged (e.g. average ToE <= 2020). A higher number of emerging stocks will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Fig 2). In most cases, the current participation of the emerging stocks in a country's total fishing revenue is less than 50%. However, in some cases emerging stocks represent 85% of total fishing revenue (e.g. Peru) or more (e.g. Marshall Islands, 92%). Only in very few cases like Brazil and Alaska's polar EEZ, no stock will emerge between 2006 and 2100.

@@

![Figure 4. Average time of emergence (ToE) per fishing entity. Fishing revenue in logarithmic scale Showing names for the top 75th percentile). Countries marked with an asterisk have at least one border.](../Figures/Fig4.png)
